---
title: "Week 09 Mid-Project Data Check-In"
subtitle: "CSC3107 Information Visualization"
author: "Team SlateBlue"
date: today
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## 1. Data Source Description

| | |
|---|---|
| **Dataset** | National Fossil Carbon Emissions 2024 v1.0 |
| **Source** | Global Carbon Project (GCB 2024) |
| **URL** | <https://globalcarbonbudget.org/gcb-2024/> |
| **File** | `National_Fossil_Carbon_Emissions_2024v1.0-1.xlsx` |

The dataset records annual fossil CO₂ emissions at the country level, covering approximately
250 countries from 1850 to 2023. Values are reported in **million tonnes of carbon per year
(MtC/yr)**. We convert to **billion tonnes of CO₂ per year (GtCO₂/yr)** using the factor
1 MtC = 3.664 MtCO₂ (as documented in the dataset).

We use the **"Territorial Emissions"** sheet, which records emissions produced within each
country's borders (excluding land-use change and consumption-based transfers).

The raw Excel file is included in `data/raw/` and the cleaned CSV outputs are in `data/cleaned/`.

## 2. Data Import and Cleaning Pipeline

### 2.1 Load Packages

```{r}
#| label: load-packages
library(tidyverse)
library(readxl)
library(scales)
```

### 2.2 Import Raw Data

```{r}
#| label: import-raw
emissions_raw <- read_excel(
  "data/raw/National_Fossil_Carbon_Emissions_2024v1.0-1.xlsx",
  sheet = "Territorial Emissions",
  skip = 11
)

# First column is the year (unnamed in Excel)
names(emissions_raw)[1] <- "year"

cat(
  "Raw dataset:",
  nrow(emissions_raw),
  "rows x",
  ncol(emissions_raw),
  "columns\n"
)
cat(
  "Year range:",
  min(emissions_raw$year, na.rm = TRUE),
  "-",
  max(emissions_raw$year, na.rm = TRUE),
  "\n"
)
```

### 2.3 Define Regions

We aggregate countries into the same five regions as the original Carbon Brief chart:
**China**, **United States (USA)**, **European Union (EU-27)**, **India**, and **Rest of World**.

```{r}
#| label: define-regions
col_names <- names(emissions_raw)

# Key country columns (verified from the dataset's column names after skip = 11)
china_col <- "China"
india_col <- "India"
usa_col <- "USA"
world_col <- "World"

# EU-27 member states
eu27_countries <- c(
  "Austria",
  "Belgium",
  "Bulgaria",
  "Croatia",
  "Cyprus",
  "Czechia",
  "Denmark",
  "Estonia",
  "Finland",
  "France",
  "Germany",
  "Greece",
  "Hungary",
  "Ireland",
  "Italy",
  "Latvia",
  "Lithuania",
  "Luxembourg",
  "Malta",
  "Netherlands",
  "Poland",
  "Portugal",
  "Romania",
  "Slovakia",
  "Slovenia",
  "Spain",
  "Sweden"
)

# Match EU-27 columns in dataset
eu27_cols <- eu27_countries[eu27_countries %in% col_names]

cat("EU-27 columns found:", length(eu27_cols), "of 27\n")
if (length(eu27_cols) < 27) {
  cat("Missing:", setdiff(eu27_countries, eu27_cols), "\n")
}
```

### 2.4 Clean and Aggregate

```{r}
#| label: clean-aggregate
# Keep only rows with valid numeric years
emissions_clean <- emissions_raw |>
  filter(!is.na(year)) |>
  mutate(year = as.numeric(year)) |>
  filter(!is.na(year))

# Conversion factor: MtC/yr -> GtCO2/yr
MTC_TO_GTCO2 <- 3.664 / 1000

# Aggregate regions
emissions_regions <- emissions_clean |>
  rowwise() |>
  mutate(
    china_gtco2 = ifelse(
      !is.na(.data[[china_col]]),
      as.numeric(.data[[china_col]]) * MTC_TO_GTCO2,
      0
    ),
    india_gtco2 = ifelse(
      !is.na(.data[[india_col]]),
      as.numeric(.data[[india_col]]) * MTC_TO_GTCO2,
      0
    ),
    usa_gtco2 = ifelse(
      !is.na(.data[[usa_col]]),
      as.numeric(.data[[usa_col]]) * MTC_TO_GTCO2,
      0
    ),
    world_gtco2 = ifelse(
      !is.na(.data[[world_col]]),
      as.numeric(.data[[world_col]]) * MTC_TO_GTCO2,
      0
    ),
    eu27_gtco2 = sum(
      across(all_of(eu27_cols), ~ ifelse(is.na(.x), 0, as.numeric(.x))),
      na.rm = TRUE
    ) *
      MTC_TO_GTCO2
  ) |>
  ungroup() |>
  mutate(
    rest_of_world_gtco2 = world_gtco2 -
      china_gtco2 -
      india_gtco2 -
      usa_gtco2 -
      eu27_gtco2
  ) |>
  select(
    year,
    china_gtco2,
    india_gtco2,
    usa_gtco2,
    eu27_gtco2,
    rest_of_world_gtco2,
    world_gtco2
  ) |>
  filter(year >= 1959, year <= 2023)

cat("Cleaned dataset:", nrow(emissions_regions), "rows (1959-2023)\n")
```

### 2.5 Pivot to Long Format

```{r}
#| label: pivot-long
emissions_long <- emissions_regions |>
  pivot_longer(
    cols = c(
      china_gtco2,
      india_gtco2,
      usa_gtco2,
      eu27_gtco2,
      rest_of_world_gtco2
    ),
    names_to = "region",
    values_to = "emissions_gtco2"
  ) |>
  mutate(
    region = case_match(
      region,
      "china_gtco2" ~ "China",
      "india_gtco2" ~ "India",
      "usa_gtco2" ~ "United States",
      "eu27_gtco2" ~ "European Union",
      "rest_of_world_gtco2" ~ "Rest of World"
    ),
    region = factor(
      region,
      levels = c(
        "China",
        "United States",
        "European Union",
        "India",
        "Rest of World"
      )
    ),
    pct_share = emissions_gtco2 / world_gtco2 * 100
  )
```

### 2.6 Save Cleaned Data

```{r}
#| label: save-cleaned
write_csv(emissions_regions, "data/cleaned/emissions_wide.csv")
write_csv(emissions_long, "data/cleaned/emissions_by_region.csv")

cat("Saved: data/cleaned/emissions_wide.csv\n")
cat("Saved: data/cleaned/emissions_by_region.csv\n")
```

### 2.7 Quick Validation

```{r}
#| label: validation
# Sanity check: 2023 global total should be ~37-38 GtCO2
total_2023 <- emissions_regions |>
  filter(year == 2023) |>
  pull(world_gtco2)
cat(
  "2023 global total:",
  round(total_2023, 2),
  "GtCO2",
  "(expected ~37.4 GtCO2)\n\n"
)

# Show latest year per region
emissions_long |>
  filter(year == 2023) |>
  arrange(desc(emissions_gtco2)) |>
  mutate(
    emissions_gtco2 = round(emissions_gtco2, 2),
    pct_share = round(pct_share, 1)
  ) |>
  select(region, emissions_gtco2, pct_share) |>
  knitr::kable(
    col.names = c("Region", "GtCO2 (2023)", "% of Global"),
    caption = "2023 emissions by region"
  )
```

**Cleaning steps summary:**

1. Imported "Territorial Emissions" sheet, skipping 11 metadata rows
2. Identified country columns: China, India, USA, and 27 EU member states
3. Converted units from MtC/yr to GtCO₂/yr (×3.664 / 1000)
4. Summed 27 EU member states into a single EU-27 aggregate
5. Computed Rest of World = World total − (China + India + USA + EU-27)
6. Filtered to 1959–2023 to match the original chart timespan
7. Pivoted to long format and computed percentage shares

## 3. Draft Design Brief

::: {.callout-note appearance="simple" icon=false}
## Design Brief

This redesign targets **general-public readers of climate news** who want to compare which
countries drive CO₂ emissions growth. The goal is to show both **total** and **per-capita
emissions** so readers can assess national responsibility on a fair basis.

**Main revisions:**

- **Sparkline summary table** for at-a-glance trend comparison across regions
- **Faceted line charts** (small multiples) replacing the stacked area chart so each region is
  readable from a shared zero baseline
- **Per-capita panel** alongside total emissions for population-adjusted context
- **Direct labels** (`ggrepel`) to eliminate legend look-up
- **Event annotations** (Paris Agreement, COVID-19) connecting trends to real-world context
- **% share view** instead of relying on stacked areas for part-to-whole comparison
- **Okabe–Ito colour palette** for colorblind accessibility
:::

## 4. Planned Design Direction

Our redesign replaces the original Carbon Brief stacked area chart with a multi-view layout.
The main panel will be a set of **faceted line charts** (small multiples)—one per region—so
that every region's trend is readable from a common zero baseline without the mental
subtraction that stacked areas require. Alongside this, we plan a **sparkline summary table**
(as suggested during the Week 5 feedback) that pairs miniature trend lines with key numbers
(latest GtCO₂, % share, 5-year % change) for quick cross-region comparison. A **percentage
share line chart** will show how each region's contribution has shifted over 65 years,
addressing the feedback that "Rest of World" could be represented as a proportion rather than
an absolute area.

We also intend to add a **per-capita emissions panel** using UN World Population Prospects
data, which we will integrate before Week 11. This addresses our proposal's critique that the
original chart overstates China's responsibility by showing only absolute totals. One open
question for the instructors: should the per-capita panel appear as a separate faceted chart
beside the total-emissions chart, or would an inset/sidebar layout work better within the
poster's 16:9 format?
